---
title: "code_for_svm"
author: "Qi Liu"
date: "8/18/2018"
output: pdf_document
---

## a)  
```{r, include = FALSE}
# Generate data
set.seed(1)
n = 40
p = 2
xpos = matrix(rnorm(n * p, mean = 0, sd = 1), n, p)
xneg = matrix(rnorm(n * p, mean = 4, sd = 1), n, p)
x = rbind(xpos, xneg)
y = matrix(c(rep(1, n), rep(-1, n)))
N = 2 * n
```  
The dual form is
$$
\min \ \left(\displaystyle\frac{1}{2}\sum_{i=1}^{n}\sum_{j=1}^{n}\alpha_{i}\alpha_{j}y_{i}y_{j}\boldsymbol{x}_{i}^{T}\boldsymbol{x}_{j}-\sum_{i=1}^{n}\alpha_{i}\right)
$$
$$
\mathrm{s.t.}
\begin{cases}
\alpha_{i} \geq 0 \\
\\
\displaystyle\sum_{i=1}^{n}\alpha_{i}y_{i}=0
\end{cases}
$$  
We rewrite $\mathrm{X}$ as 
$$
\mathrm{X}=
\left[ \begin{array}{c}
\boldsymbol{x}_{1}^{\mathrm{T}} \\
\boldsymbol{x}_{2}^{\mathrm{T}} \\
\vdots\\
\boldsymbol{x}_{N}^{\mathrm{T}} \\
\end{array} \right],
$$  
where $N$ is the total number of sample points.

Now rewrite this dual form with matrix and vector, we can get  
$$
\min \left(\displaystyle\frac{1}{2}\boldsymbol{\alpha}^{\mathrm{T}}\mathrm{Q}^{\mathrm{T}}\mathrm{Q}\boldsymbol{\alpha}-\boldsymbol{d}^{\mathrm{T}}\boldsymbol{\alpha}\right)
$$
where,
$$
\quad  \boldsymbol{\alpha} = \left[\alpha_{1}, \alpha_{2},\cdots ,\alpha_{N}\right]^{\mathrm{T}}, \boldsymbol{d}=[1, 1, \cdots, 1]^{\mathrm{T}},\\
\mathrm{Q}=\left[y_{1}\boldsymbol{x}_{1},\  y_{2}\boldsymbol{x}_{2},\ \cdots,\ y_{N}\boldsymbol{x}_{N}\right]
$$  
so $\mathrm{D}=\mathrm{Q}^{\mathrm{T}}\mathrm{Q}$.
And the matrices in constrains part are 
$$
\mathrm{A}^{\mathrm{T}}=
\left[ \begin{array}{cccc}
y_{1} \quad y_{2} \quad \cdots \quad y_{n} \\
1     \quad  0    \ \quad \cdots \quad  0    \\
0     \quad  1    \ \quad \cdots \quad  0    \\
\vdots \quad \vdots \quad \vdots \quad \vdots \\
0     \quad  0    \ \quad \cdots \quad  1    \\
\end{array} \right]
and \ \ \boldsymbol{b}_{0}=[0, 0, \cdots, 0]^{\mathrm{T}}.
$$
Then use `quadprog` to solve this dual form.  
```{r Solve Dual Form}
library(quadprog)

# Helper function for generating Q
helper_fun = function(i) {
  y[i] * (t(x)[, i])
}

# Represent the dual form using matrix
Q = sapply(1:N, helper_fun)
D = t(Q) %*% Q
d = matrix(1, nrow = N, ncol = 1)
A = t(rbind(matrix(y, nrow = 1, ncol = N), diag(nrow = N)))
b_0 = rbind(matrix(0, nrow = 1, ncol = 1), matrix(0, nrow = N, ncol = 1))
eps = 5e-4

# Calculate the dual form about maximization over alpha
soln = solve.QP(Dmat = D + eps * diag(N), dvec = d, 
                Amat = A, bvec = b_0, 
                meq = 1, factorized = FALSE)
qpsoln = matrix(soln$solution, nrow = N)

# Find the nonzero part of solution
nonzero = abs(qpsoln) > 1e-5

# Helper function for w
w_helper = function(i) {
  qpsoln[i] * y[i] * x[i, ]
}
w = rowSums(sapply(1:N, w_helper))
b = -(max(x[y == -1, ] %*% w) + min(x[y == 1, ] %*% w)) / 2
```  
The support vector is 
```{r}
(support_vectors = x[which(nonzero), ])
```  
and the decision line is `r w[1]`$x_{1}$ `r w[2]`$x_{2}+$ `r b`$=0$.

Now we plot the solution as following,  
```{r, echo = FALSE}
library(ggplot2)
ggplot(data.frame(x, y), aes(x = X1, y = X2)) + 
  ggtitle("SVM Dual From") + 
  geom_point(aes(colour = factor(y))) + 
  geom_point(data = data.frame(x[which(nonzero), ]), aes(x = X1, y = X2), 
             shape = 1, size = 5) + 
  geom_abline(intercept = -b / w[2], slope = -w[1] / w[2]) + 
  geom_abline(intercept = -(b - 1) / w[2], slope = -w[1] / w[2], 
              linetype = "dashed") + 
  geom_abline(intercept = -(b + 1) / w[2], slope = -w[1] / w[2], 
              linetype = "dashed") + 
  theme_bw()
```
Next, we compare our conclusions with the SVM model.
```{r Check with "svm"}
library(e1071)
svm.fit = svm(y ~ ., data = data.frame(x, y),
              type = "C-classification",
              kernel = "linear",
              scale = FALSE,
              cost = 10000)
w = t(svm.fit$coefs) %*% svm.fit$SV
b = -svm.fit$rho 
```  
Now plot the results from `e1071`.
```{r, echo = FALSE}
# Plot
ggplot(data.frame(x, y), aes(x = X1, y = X2)) + 
  ggtitle("SVM with e1071") + 
  geom_point(aes(colour = factor(y))) + 
  geom_point(data = data.frame(x[svm.fit$index, ]), aes(x = X1, y = X2), 
             shape = 1, size = 5) + 
  geom_abline(intercept = -b / w[1, 2], slope = -w[1, 1] / w[1, 2]) + 
  geom_abline(intercept = -(b - 1) / w[1, 2], slope = -w[1, 1] / w[1, 2], 
              linetype = "dashed") + 
  geom_abline(intercept = -(b + 1) / w[1, 2], slope = -w[1, 1] / w[1, 2], 
              linetype = "dashed") + 
  theme_bw()
```  

## b)  
First, generate the nonseparable data. 
```{r Generate nonseparable data}
n = 10
p = 2
set.seed(70)
xpos = matrix(rnorm(n * p, mean = 0, sd = 1), n, p)
xneg = matrix(rnorm(n * p, mean = 1.5, sd = 1), n, p)
x = rbind(xpos, xneg)
y = matrix(c(rep(1, n), rep(-1, n)))
N = 2 * n
```  
Now we have generated the nonseparable data and plot them.  
```{r, echo = FALSE}
ggplot(data.frame(x, y), aes(x = X1, y = X2)) + 
  ggtitle("Nonseparable Data") + 
  geom_point(aes(colour = factor(y))) + 
  theme_bw()
```  
$\mathrm{D}$ and $\boldsymbol{d}$ are the same as that in the separable case, and matrices in the constrains are,  
$$
\mathrm{A} = 
\left[\begin{array}{ccccc}
y_{1} \quad y_{2} \quad y_{3} \quad \cdots \quad y_{N} \\
1     \quad   0   \quad   0   \quad   \cdots  \quad 0 \\
0     \quad   1   \quad   0   \quad   \cdots  \quad 0 \\
0     \quad   0   \quad   1   \quad   \cdots  \quad 0 \\
\vdots \\
0     \quad   0   \quad   0   \quad   \cdots  \quad 1 \\
-1     \quad   0   \quad   0   \quad   \cdots  \quad 0 \\
0     \quad   -1   \quad   0   \quad   \cdots  \quad 0 \\
0     \quad   0   \quad   -1   \quad   \cdots  \quad 0 \\
\vdots \\
0     \quad   0   \quad   0   \quad   \cdots  \quad -1 \\
\end{array}\right]_{(2N+1) \times N}
$$  
and, 
$$
\boldsymbol{b}_{0}=
\left[\begin{array}{c}
0 \\
0 \\
\vdots\\
0
\end{array}\right]_{2N+1}
$$

Use `quadprog` to solve the dual form. The process is similar to that in part a), so the code is not shown here. 
```{r, echo = FALSE}
# Represent the dual form with matrix
# Helper function for generating Q
helper_fun = function(i) {
  y[i] * (t(x)[, i])
}
Q = sapply(1:N, helper_fun)
D = t(Q) %*% Q
d = matrix(1, nrow = N, ncol = 1)
A = t(rbind(matrix(y, nrow = 1, ncol = N),
            diag(N),
            -diag(N)))
# Cost term
c = 1
b_0 = rbind(matrix(0, nrow = 1, ncol = 1),
            matrix(0, nrow = N, ncol = 1),
            matrix(-c, nrow = N, ncol = 1))
eps = 5e-4

# Calculate the dual form about maximization over alpha
soln = solve.QP(Dmat = D + eps * diag(N), dvec = d, 
                Amat = A, bvec = b_0, 
                meq = 1, factorized = FALSE)
qpsoln = matrix(soln$solution, nrow = N)

# Find the nonzero part of solution
nonzero = abs(qpsoln) > 1e-5

# Helper function for w
w_helper = function(i) {
  qpsoln[i] * y[i] * x[i, ]
}
nonzero = abs(qpsoln) > 1e-5
w = rowSums(sapply(1:N, w_helper))

margin_vec = (abs(qpsoln) > 1e-5) & (abs(qpsoln) < c - 1e-5)

# Helper function for b
b_helper = function(i) {
  x[i, ] %*% w - y[i]
}
b = -mean(sapply(which(margin_vec), b_helper))
```  
The support vectors are  
```{r}
(x[which(nonzero), ])
```  
Now we plot the solution.  
```{r, echo = FALSE}
ggplot(data.frame(x, y), aes(x = X1, y = X2)) + 
  ggtitle("Nonseparable SVM Dual Form") + 
  geom_point(aes(colour = factor(y))) + 
  geom_point(data = data.frame(x[which(nonzero), ]), aes(x = X1, y = X2), 
             shape = 1, size = 5) + 
  geom_abline(intercept = -b / w[2], slope = -w[1] / w[2]) + 
  geom_abline(intercept = -(b - 1) / w[2], slope = -w[1] / w[2], 
              linetype = "dashed") + 
  geom_abline(intercept = -(b + 1) / w[2], slope = -w[1] / w[2], 
              linetype = "dashed") + 
  theme_bw()
```  

Now use `e1071` package to solve this SVM for comparison.  
```{r}
svm_fit = svm(y ~ ., data = data.frame(x, y), 
              type = "C-classification",
              kernel = "linear",
              scale = FALSE,
              cost = 1)
w = t(svm_fit$coefs) %*% svm_fit$SV
b = -svm_fit$rho
```  
Plot the result.  
```{r, echo = FALSE}
# Plot
ggplot(data.frame(x, y), aes(x = X1, y = X2)) + 
  ggtitle("Nonseparable SVM with e1071") + 
  geom_point(aes(colour = factor(y))) + 
  geom_point(data = data.frame(x[svm_fit$index, ]), aes(x = X1, y = X2),
             shape = 1, size = 5) + 
  geom_abline(intercept = -b / w[1, 2], slope = -w[1, 1] / w[1, 2]) + 
  geom_abline(intercept = -(b - 1) / w[1, 2], slope = -w[1, 1] / w[1, 2], 
              linetype = "dashed") + 
  geom_abline(intercept = -(b + 1) / w[1, 2], slope = -w[1, 1] / w[1, 2], 
              linetype = "dashed") + 
  theme_bw()
```

## c)  
The new deficion function is 
$$
\begin{aligned}
f(\boldsymbol{x}) 
&= \boldsymbol{x}^{\mathrm{T}}\boldsymbol{\beta}+\beta_{0}\\
&=\boldsymbol{x}^{\mathrm{T}}\sum_{i=1}^{N}\alpha_{i}y_{i}\boldsymbol{x}_{i} + \beta_{0} \\
&=\sum_{i=1}^{N}\alpha_{i}y_{i}(\boldsymbol{x}^{\mathrm{T}}\boldsymbol{x}_{i})+\beta_{0} \\
&=\sum_{i=1}^{N}\alpha_{i}y_{i}(\Phi(\boldsymbol{x})^{\mathrm{T}}\Phi(\boldsymbol{x}_{i}))+\beta_{0}
\end{aligned}
$$  
Using the relationship between kernel function and basis expansions, 
$$
f(\boldsymbol{x})=\sum_{i=1}^{N}\alpha_{i}y_{i}\mathrm{K}(\boldsymbol{x}, \boldsymbol{x}_{i})+\beta_{0}
$$  

## d)  
The optimization part of the dual problem can be written as $\mathrm{min}\quad \left(\displaystyle\frac{1}{2}\displaystyle\sum_{i=1}^{N}\displaystyle\sum_{j=1}^{N}\alpha_{i}\alpha_{j}y_{i}y_{j}\mathrm{K}(\boldsymbol{x}_{i}, \boldsymbol{x}_{j})-\displaystyle\sum_{i=1}^{N}\alpha\right)$. Define a new vector $\boldsymbol{\mathrm{Q}}=\left[y_{1}\Phi(\boldsymbol{x}_{1}),\cdots, y_{N}\Phi(\boldsymbol{x}_{N})\right]$, then
$$
\boldsymbol{\mathrm{D}}=\boldsymbol{\mathrm{Q}}^{\mathrm{T}}\boldsymbol{\mathrm{Q}}=\left[y_{i}y_{j}\mathrm{K}(\boldsymbol{x}_{i}, \boldsymbol{x}_{j})\right]_{i, j=1,\cdots, N}
$$
$\boldsymbol{\mathrm{A}}$, $d$ and $b_{0}$ are the same as part b.